Using GPT-4 to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2025-01-07 manuscript.qmd
edit manuscript

---
title: Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 0000-0002-8669-3076
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 0000-0001-5949-2191
    corresponding: false
    email: brian_schilder@alumni.brown.edu
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 0000-0002-6807-3180
    email: n.skene@imperial.ac.uk
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - Ontology
  - Human Phenotype Ontology
  - Large Language Model
  - GPT-4
  - Artificial intelligence
  - Generative AI
  - Rare disease 
  - Automation
  - Medicine
bibliography: references.bib
date: last-modified
citation: 
  container-title: medRxiv
---

@@

There are thousands of human phenotypes which are linked to genetic variation. These range from the benign (white eyelashes) to the deadly (respiratory failure). The Human Phenotype Ontology has categorised all human phenotypic variation into a unified framework that defines the relationships between them (e.g. missing arms and missing legs are both abnormalities of the limb). This has made it possible to perform phenome-wide analyses, e.g. to prioritise which make the best candidates for gene therapies. However, there is currently limited metadata describing the clinical characteristics / severity of these phenotypes. With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO. Phenotypic severity was defined using a set of clinical characteristics and their frequency of occurrence. First, we benchmarked the generative LLM clinical characteristic annotations against ground-truth labels within the HPO (e.g. phenotypes in the 'Cancer' HPO branch were annotated as causing cancer by GPT-4). True positive recall rates across different clinical characteristics ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the clinical characteristic and the frequency of its occurrence. These clinical characteristic severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.

@@

This high level of recall underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information. However, we acknowledge that the number of testable true positive phenotypes for some of these categories are low, especially `r shQuote(names(checks$checkable_count)[checks$checkable_count==min(checks$checkable_count)])` for which there is only `r min(checks$checkable_count)` phenotype in the HPO (after excluding terms pertaining to colour or night blindness). Furthermore, some of the true positive phenotypes are lexically similar to the name of the clinical characteristic itself. In these cases, annotating 'Severe intellectual disability' as always causing intellectual disability is a relatively trivial task. Nevertheless, even these scenarios provide a clear and interpretable benchmark for the model's performance. In addition, were numerous phenotypes with lexically non-obvious relationships to the clinical characteristic that were annotated correctly by GPT-4. For example, 'Molar tooth sign on MRI' (a neurodevelopmental pathology observed in radiological scans) was correctly annotated as causing intellectual disability.

@@

While individual annotations are informative, we wanted to be able to distil the severity of each phenotype into a single score. Quantifying the overall severity of phenotypes can have important implications for diagnosis, prognosis, and treatment. It may also guide the prioritisation of gene therapy trials for phenotypes with the most severe clinical characteristics and thus the most urgent need. Importantly, the values reflected the severity of each clinical characteristic based on both the type of characteristic itself and its frequency within a particular phenotype. For instance, a phenotype always causing death would have a higher multiplied value than a phenotype often causing reduced fertility (see @tbl-metric-weights). First, we created a dictionary to map each clinical characteristic (e.g. blindness) and its frequency (always, often, rarely, never) to numeric values from 0-3. Then, the clinical characteristic values were multiplied by weights. Next, we computed an average score for each phenotype by aggregating the multiplied values across all clinical characteristics and then calculating the mean. This was then normalised by the theoretical maximum severity score, so that all phenotypes were on a 0-100 severity scale (where 100 is the most severe phenotype possible). This average normalised score represents the overall severity of the phenotype based on the severity of its individual clinical characteristics.

@@

```{r fig-top-phenos}
#| label: fig-top-phenos
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital (**a**-**b**) or rarely/never congenital (**c**-**d**). Continuous severity scores are shown as bars (**b**,**d**) and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights. The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axes show each of the clinical characteristics. All data for this figure, as well as justifications for each annotation, can be found in @tbl-annotations.
#| fig-height: 10
#| fig-width: 13

@@

We found that some clinical characteristic severity metrics were correlated with one another, with a mean Pearson correlation of `r round(mean(cor_metrics$r[-nrow(cor_metrics$r),-ncol(cor_metrics$r)]),2)` across all individual metrics (see @fig-metric-corplot). In particular, blindness and sensory impairment were highly correlated with one another (r=`r round(cor_metrics$r["blindness","sensory_impairments"],2)`, p=`r round(cor_metrics$P["blindness","sensory_impairments"],4)`). Some metrics drove the composite severity score more than other, which is a reflection of both our per-metric weighting scheme, response type frequencies, and the correlation structure between metrics. Overall, `r gsub("_"," ",names(cor_composite[1]))` seemed to be the strongest driver of the composite severity score with a Pearson correlation of `r mean(cor_composite[1],2)`, followed by `r gsub("_"," ",names(cor_composite[2]))` (r=`r round(cor_composite[2],2)`) and `r gsub("_"," ",names(cor_composite[3]))` (r=`r round(cor_composite[3],2)`).

@@

{{< pagebreak >}

@@

```{r fig-severity-histo}
#| label: fig-severity-histo
#| fig-cap: Distribution of the composite GPT-4 severity score of the severity scores for all HPO terms.
#| fig-height: 8
#| fig-width: 10

>>>>

---
title: Using GPT-4 to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 0000-0002-8669-3076
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 0000-0001-5949-2191
    corresponding: false
    email: brian_schilder@alumni.brown.edu
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 0000-0002-6807-3180
    email: n.skene@imperial.ac.uk
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - Ontology
  - Human Phenotype Ontology
  - Large Language Model
  - GPT-4
  - Artificial intelligence
  - Generative AI
  - Rare disease 
  - Automation
  - Medicine
bibliography: references.bib
date: last-modified
citation: 
  container-title: medRxiv
---

@@

There are thousands of human phenotypes which are linked to genetic variation. These range from the benign (white eyelashes) to the deadly (respiratory failure). The Human Phenotype Ontology has categorised a vast array of human phenotypic variation into a unified framework that defines the relationships between them (e.g. missing arms and missing legs are both abnormalities of the limb). This has made it possible to perform phenome-wide analyses, e.g. to prioritise which make the best candidates for gene therapies. However, there is currently limited metadata describing the clinical characteristics / severity of these phenotypes. With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO. Phenotypic severity was defined using a set of clinical characteristics and their frequency of occurrence. First, we benchmarked the generative LLM clinical characteristic annotations against ground-truth labels within the HPO (e.g. phenotypes in the 'Cancer' HPO branch were annotated as causing cancer by GPT-4). True positive recall rates across different clinical characteristics ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the clinical characteristic and the frequency of its occurrence. These clinical characteristic severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.

@@

This high level of recall underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information. However, we acknowledge that the number of testable true positive phenotypes for some of these categories are low, especially `r shQuote(names(checks$checkable_count)[checks$checkable_count==min(checks$checkable_count)])` for which there is only `r min(checks$checkable_count)` phenotype in the HPO (after excluding terms pertaining to colour or night blindness, as these are not typically classified as causes of complete blindness). Furthermore, some of the true positive phenotypes are lexically similar to the name of the clinical characteristic itself. In these cases, annotating 'Severe intellectual disability' as always causing intellectual disability is a relatively trivial task. Nevertheless, even these scenarios provide a clear and interpretable benchmark for the model's performance. In addition, were numerous phenotypes with lexically non-obvious relationships to the clinical characteristic that were annotated correctly by GPT-4. For example, 'Molar tooth sign on MRI' (a neurodevelopmental pathology observed in radiological scans) was correctly annotated as causing intellectual disability.

@@

While individual annotations are informative, we wanted to be able to distil the severity of each phenotype into a single score. Quantifying the overall severity of phenotypes can have important implications for diagnosis, prognosis, and treatment. It may also guide the prioritisation of gene therapy trials for phenotypes with the most severe clinical characteristics and thus the most urgent need. Importantly, the values reflected the severity of each clinical characteristic based on both the type of characteristic itself and its frequency within a particular phenotype. For instance, a phenotype always causing death would have a higher multiplied value than a phenotype often causing reduced fertility (see @tbl-metric-weights). First, we created a dictionary to map each clinical characteristic (e.g. blindness) and its frequency (always, often, rarely, never) to numeric values from 0-3. Then, the clinical characteristic values were multiplied by weights. Next, we computed an average score for each phenotype by aggregating the multiplied values across all clinical characteristics and then calculating the mean. This was then normalised by the theoretical maximum severity score, so that all phenotypes were on a 0-100 severity scale (where 100 is the most severe phenotype possible). This average normalised score represents the overall severity of the phenotype based on the severity of its individual clinical characteristics. @fig-severity-histo illustrates the distribution of severity scores across HPO terms by branch, with 'Abnormal cellular phenotype' being the branch that contains phenotypes with the highest severity scores on average.

```{r fig-severity-histo}
#| label: fig-severity-histo
#| fig-cap: Distribution of composite GPT-4 severity scores across HPO terms, grouped by branch. The figure shows the distribution of phenotypes for each severity score (ranging from 0 to 100) across HPO branches. The vertical line indicates the average severity score for each branch. The branch 'Abnormal cellular phenotype' contains the highest concentration of terms associated with the most severe clinical characteristics.
#| fig-height: 8
#| fig-width: 10

@@

```{r fig-top-phenos}
#| label: fig-top-phenos
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital **(a-b)** or rarely/never congenital **(c-d)**. Continuous severity scores are shown as bars **(b,d)** and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights. The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axes show each of the clinical characteristics. All data for this figure, as well as justifications for each annotation, can be found in @tbl-annotations.
#| fig-height: 10
#| fig-width: 13

@@

We found that some clinical characteristic severity metrics were correlated with one another, with a mean Pearson correlation of `r round(mean(cor_metrics$r[-nrow(cor_metrics$r),-ncol(cor_metrics$r)]),2)` across all individual metrics (see @fig-metric-corplot). In particular, blindness and sensory impairment were highly correlated with one another (r=`r round(cor_metrics$r["blindness","sensory_impairments"],2)`, p=`r round(cor_metrics$P["blindness","sensory_impairments"],4)`). Some metrics drove the composite severity score more than other, which is a reflection of both our per-metric weighting scheme, response type frequencies, and the correlation structure between metrics. Overall, `r gsub("_"," ",names(cor_composite[1]))` seemed to be the strongest driver of the composite severity score with a Pearson correlation of `r mean(cor_composite[1],2)`, followed by `r gsub("_"," ",names(cor_composite[2]))` (r=`r round(cor_composite[2],2)`) and `r gsub("_"," ",names(cor_composite[3]))` (r=`r round(cor_composite[3],2)`). This is driven by the greater number of phenotypes 'always' causing impaired mobility (422), compared with, for instance, death (66), and intellectual disability (64).

@@

@@
